---
title: "Linear_Regression"
author: "Yi Hung Chen, Kyriakos Papadopoulos"
date: "21/09/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear_Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
### Description
This package is used to conduct a simple regression analysis using iris dataset and some methods have been made to present stats from the linear regression model.

## Created a linear regression model

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Lab4)
mod_object <- linreg(Petal.Length~Species, data = iris)
```

### Print()
> *should print out the coefficients and coefficient names*
```{r, eval=TRUE, include=TRUE}
mod_object$print()
```
### Plot()
```{r}
list_of_plots = mod_object$plot()
```

- 1st plot
```{r, eval=TRUE, include=TRUE}
list_of_plots[1]
```

- 2nd plot
```{r, eval=TRUE, include=TRUE}
list_of_plots[2]
```

### resid()
> *Returns the vector of residuals*
```{r, eval=TRUE, include=TRUE}
mod_object$resid()[1:4]
```

### pred()
> *Returns the predicted values*
```{r, eval=TRUE, include=TRUE, output.lines=4}
mod_object$pred()[1:4]
```

### coef()
> *Returns the coefficients*
```{r, eval=TRUE, include=TRUE}
mod_object$coef()
```

### summary()
> *Presents the coefficients with their standard error, t-value and p-value as well as the estimate of ˆσ and the degrees
of freedom in the model.*
```{r, eval=TRUE, include=TRUE}
mod_object$summary()
```
