---
title: "Lab6"
author: "Yi Hung Chen, Kyriakos Papadopoulos"
date: "03/10/2022"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lab6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

### Description

This Package contains 3 different algorithms for solving the **knapsack problem**. The 3 algorithms that we use are

1.  Brute force search
2.  Dynamic Programming
3.  Greedy Heuristic

```{r, include = FALSE}
RNGversion(min(as.character(getRversion()),"3.5.3"))

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(42, kind = "Mersenne-Twister", normal.kind = "Inversion")
n <- 2000
knapsack_objects <-
data.frame(
w=sample(1:4000, size = n, replace = TRUE),
v=runif(n = n, 0, 10000)
)
```

```{r setup}
library(Lab6)
```

### brute_force_knapsack(x,W,parallel = FALSE)

> The complexity of the algorithm is 2\^(n) as avaluates all possible combinations

-   x -\> The matrix which contains the weight and value of each item
-   W -\> The total weight that knapsack fit
-   parallel -\> By default is false and it gives the opportunity to the user to select if he/she wants to run the algorithm in parallel.

```{r, eval=TRUE, include=TRUE}
brute_force_knapsack(x = knapsack_objects[1:8,], W = 3500)
brute_force_knapsack(x = knapsack_objects[1:12,], W = 3500)
brute_force_knapsack(x = knapsack_objects[1:8,], W = 2000)
brute_force_knapsack(x = knapsack_objects[1:12,], W = 2000)
```

The time that it takes to run the algorithm for n=16 is 144.68s 

(Since the run time is high, we do not run the code in real time when vignette is created)

### knapsack_dynamic(x,W)

> This algorithm iterates all possible values of W. It will run in Wn.

-   x -\> The matrix which contains the weight and value of each item
-   W -\> The total weight that knapsack fit

```{r, eval=TRUE, include=TRUE}
knapsack_dynamic(x = knapsack_objects[1:8,], W = 3500)
knapsack_dynamic(x = knapsack_objects[1:12,], W = 3500)
knapsack_dynamic(x = knapsack_objects[1:8,], W = 2000)
knapsack_dynamic(x = knapsack_objects[1:12,], W = 2000)
```

The time that it takes to run the algorithm for n=500 is 2.28 seconds.

### greedy_knapsack(x,W)
> This algorithm will not give an exact result but it's complexity is nlogn

-   x -\> The matrix which contains the weight and value of each item
-   W -\> The total weight that knapsack fit

```{r, eval=TRUE, include=TRUE}
greedy_knapsack(x = knapsack_objects[1:800,], W = 3500)
greedy_knapsack(x = knapsack_objects[1:1200,], W = 2000)
```

The time that it takes to run the algorithm for n=1000000 is 0.09s seconds.


### Profiling
>After using profvis to profile our brute_force algorithm, we found out using get_value and get_indexes in brute_force is one of the bottleneck, so we did some change in get_value & get_indexes function

We remove the for loop in both functions, and replace it with "rowSums"
```{r}
get_indexes<- function(vector,data){
  indexes <- which(vector == 1)
  
  sum <- 0
  if (exists("indexes"))
  {
    # for (i in indexes){
    #   sum <- sum + data[i, 1]
    # }
    sum <- sum(rowSums(as.matrix(data[indexes, 1])))

  }
  return(sum)
}
```

```{r}
get_value <- function(vector,data){
  indexes <- which(vector == 1)
  sum <- 0
  if (exists("indexes"))
  {
    #for (i in indexes){
    # sum <- sum + data[i, 2]
    #}
    sum <- sum(rowSums(as.matrix(data[indexes, 2])))
  }
  return(sum)
}
```

We use knapsack_objects[1:12,] to test our result and we got 8% speed improvement 
(Originally 2.33 secs and After 2.14 secs)

### Parallelizing
>We using parallelizing to create the dataframe with all combinations in brute force algorithm. 

```{r, eval=FALSE}
  if (parallel == TRUE) {
    cores <- parallel::detectCores()
    cl <- parallel::makeCluster(cores, type = "PSOCK")

    n <- nrow(x)
    l <- rep(list(0:1), n)
    
    combinations<- parallel::parLapply(cl, 1:(2^n), function(x) as.integer(intToBits(x)))
    parallel::stopCluster(cl)
   }
```
  
However, the improvement is not noticeable. Using system.time to check with 12 elements. We notice that we got slight improvement in User instruction but lose time in System instruction. 